The Calculus of Linear Constructions

### 2022-01-21 writing/draft7/draft7.tex
draft

\begin{document}
\begin{abstract}
  The Calculus of Linear Constructions (CLC) is an extension of the Calculus of Constructions (CC) with linear types. Specifically, CLC extends the predicative CC$\omega$ with a hierarchy of linear universes that precisely controls the weakening and contraction of its term level inhabitants. We study the meta-theory of CLC, showing that it is a sound logical framework for reasoning about resource. We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Through examples, we demonstrate that CILC facilitates correct by construction imperative programming and lightweight protocol enforcement. We have formalized and proven correct all major results in the Coq Proof Assistant.
\end{abstract}
\maketitle

\section{Introduction}
The Calculus of Constructions (CC) is a dependent type theory introduced by Coquand and Huet in their landmark work \cite{cc}. In CC, types can depend on terms, allowing one to write precise specifications as types. Today, CC and its extensions CIC \cite{cic} and ECC \cite{ecc} lie at the core of popular proof assistants such as Coq \cite{coq}, Agda \cite{agda}, Lean \cite{lean} and others. These theorem provers have found great success in the fields of software verification and constructive mathematics.

However, due to its origins as a logical framework for constructive mathematics, it can be inconvenient for CC to encode and reason about resource. Users of proof assistants based on CC often need to embed external logics such as Separation Logic to provide domain-specific reasoning principles for dealing with resource. We propose an alternative solution: extend CC with linear types.

This paper presents a new linear dependent type system --- the Calculus of Linear Constructions (CLC). CLC extends the predicative CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that every resource is used exactly once.

Preexisting approaches to integrating linear types and dependent types often relied on the ! exponential or separate typing judgments for linear and non-linear types. The universe sorts of CLC are enough to completely distinguish between linear and non-linear types within a single typing judgment. As a result, CLC falls closer to its CC roots than prior works in both syntax and semantics.

We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Standard Linear Logic connectives such as $\otimes$ and $\oplus$ are definable using this generalized mechanism in a straightforward way. Connectives that mix linear and non-linear types can be constructed just as easily using CILC.

Through examples in Section \ref{examples}, we show some applications of CILC such as construction of safe random access array from the first principles, protocol enforcement through inductively defined session types and in-place update of data structures using continuations.

We have formalized all major results in Coq and implemented a prototype in OCaml. Additional examples of applications such as provable effect laws and quantitative reasoning are included with our implementation.

\vspace{4pt}
\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item First, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types, enabling direct and precise reasoning about resource.
  \item Second, we study the meta-theory of CLC, showing that it exhibits many desirable properties such as confluence, subject reduction and logical consistency.
  \item Furthermore, we extend CLC with rules to define inductive linear types, forming the Calculus of Inductive Linear Constructions (CILC). This is the first complete presentation of a generalized mechanism for defining inductive linear types.
  \item All major results have been proven correct in the Coq. Our development is the first machine checked formalization of a linear dependent type theory. We have also implemented a prototype of CILC in OCaml.
\end{itemize}

@@

The syntax of the core type theory is presented in Figure \ref{syntax1}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $i$ to quantify over universe levels $0, 1, 2, ...$. $U_i$ and $L_i$ are the universes at level $i$ of non-linear and linear types respectively.
\begin{figure}[h]
  \caption{Correspondence of CLC types and MELL implications}
  \Description{}
  \begin{align}
    (\_ \utype A) \rightarrow B \quad & \Leftrightarrow \quad !(!A \multimap B) \\
    (\_ \ltype A) \rightarrow B \quad & \Leftrightarrow \quad !(A \multimap B)  \\
    (\_ \utype A) \multimap B \quad   & \Leftrightarrow \quad !A \multimap B    \\
    (\_ \ltype A) \multimap B \quad   & \Leftrightarrow \quad A \multimap B
  \end{align}
  \label{correspondence}
\end{figure}

A clear departure of our language from standard presentations of both linear type theory and dependent type theory is the presence of two function types: $(x :_s A) \rightarrow B$ and $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic implicitly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theories. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we have limited ! to only be canonically usable. This allows us to derive the substitution lemma and construct a direct modeling of CLC in CC$\omega$ without requiring any additional machinery for manipulating exponential.

Figure \ref{correspondence} illustrates the correspondence between CLC functions and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (3) if the co-domain $B$ is dependent on arguments of domain $A$. We will discuss function type formation in Section \ref{tyformation} in greater detail.

@@

Figure \ref{universe} illustrates the structure of our universe hierarchy. Each linear universe $L_i$ has $U_{i+1}$ as its type, allowing functions to freely quantify over linear \textit{types}. However, $L_i$ cumulates to $L_{i+1}$. These two parallel threads of cumulativity prevent linear types from being transported to the non-linear universes and subsequently losing track of linearity.

@@

An auxiliary judgment $\pure{\Gamma}$ is defined to assert that a context $\Gamma$ does not contain linear variables. In other words, all variables found in $\pure{\Gamma}$ are annotated of the form $x \utype A$. The full rules for structural judgments are presented in Figure \ref{structural}.

@@

\subsection{Type Formation} \label{tyformation}
The rules for forming types are presented in Figure \ref{type}. In CLC, we forbid types from depending on linear terms similarly to \cite{llf,neel15} for the same reason of avoiding philosophical troubles.

The axiom rules \rname{U-Axiom} and \rname{L-Axiom} are similar to standard CC$\omega$, the main difference being the extra side-condition of judgment $\pure{\Gamma}$. In most presentations of dependent type theories without linear types, the universe axioms are derivable under any well-formed context $\Gamma$. Variables not pertaining to actual proofs could be introduced this way, thus giving rise to the weakening rule. To support linear types, we must restrict weakening to non-linear variables. This justifies the restriction of $\Gamma$ to contain only non-linear variables for \rname{U-Axiom} and \rname{L-Axiom}. From \rname{L-Axiom} we can see that the universe of linear types $L_i$ is an inhabitant of $U_{i+1}$. This is inspired by Krishnaswami et al.'s treatment of linear universes \cite{neel15}, where linear \textit{types} themselves can be used unrestrictedly.

@@

\subsection{Equality and Reduction} \label{reduction}
The operational semantics and $\beta$-equality of CLC terms are presented in Figure \ref{red}, all of which are entirely standard.

As we have discussed previously, the elimination of the explicit ! exponential allows CLC to maintain the standard operational semantics of CC$\omega$, whose $\beta$-reductions are very well behaved.

@@

\subsection{Meta Theory} \label{meta}
In this section, we focus our discussion on the properties of CLC. First, we show the type soundness of CLC through the \textit{subject reduction} theorem. Next, through the \textit{linearity} theorem we illustrate how CLC can effectively control usage of resource. Furthermore, we demonstrate that the \textit{promotion} and \textit{dereliction} rules can be encoded as $\eta$-expansions. Finally, we construct a reduction preserving erasure function that maps well-typed CLC terms to well-typed CC$\omega$ terms, proving that CLC is strongly normalizing.

All proofs have been formalized in Coq with help from the Autosubst \cite{autosubst} library. The Coq development is publicly available on the author's Github repository. To the best of our knowledge, this is the first machine checked formalization of a linear dependently type theory. We give a hand written version of the proof in an accompanying technical report.

@@

The proof of confluence is entirely standard using parallel reduction technique \cite{takahashi}. The presence of linear types does not pose any complications as reductions are untyped.

@@

\subsubsection{Substitution} \label{subst}
Although the substitution lemma is regarded as a boring and bureaucratic result, it is surprisingly hard to design linear typed languages where the substitution lemma is admissible. Much of the difficulty arises during the substitution of arguments containing ! exponential. Perhaps the most famous work detailing the issues of substitution is due to Wadler \cite{substitute}. He defines intricate syntax and semantics for the unboxing of ! terms, solving the lack of substitute in Abramsky's term calculus.

@@

Notice in Theorem \ref{nlsubst} of non-linear substitutions, there exists an extra condition $\pure{\Gamma_2}$ which Theorem \ref{lsubst} of linear substitutions lacks. This condition ensures that linear variables do not occur within the substituted term, thus cannot be duplicated by substitution.

@@

\subsubsection{Strong Normalization}
The strong normalization theorem of CLC is proven by construction of a typing and reduction preserving erasure function from CLC to CC$\omega$. We assume familiarity with CC$\omega$ syntax here and define the erasure function as follows.

@@

For the following theorems, we refer to the reductions in CLC as $\step_{\scriptscriptstyle \text{CLC}}$, and the reductions in CC$\omega$ as $\step_{\scriptscriptstyle \text{CC$\omega$}}$.

\begin{theorem} \label{preserve}
  Reduction Erasure. For any CLC terms $m$ and $n$, if there is $m \step_{\scriptscriptstyle \text{CLC}} n$, then there is $\erase{m} \step_{\scriptscriptstyle \text{CC$\omega$}} \erase{n}$.
\end{theorem}

\begin{theorem} \label{embedding}
  Embedding. If $\Gamma \vdash m \stype{s} A$ is a valid judgment in CLC, then $\erase{\Gamma} \vdash \erase{m} : \erase{A}$ is a valid judgment in CC$\omega$.
\end{theorem}

If there exists some well-typed CLC term with an infinite sequence of reductions, erasure will embed this term into a well typed CC$\omega$ term along with its infinite sequence of reductions by virtue of theorems \ref{preserve} and \ref{embedding}. This is contradictory to the strong normalization property of CC$\omega$ proven through the Girard-Tait method \cite{ecc}, so this hypothetical term does not exist in CLC.

@@

\subsubsection{Compatibility}
To show that CLC is compatible with the predicative CC$\omega$, we construct a function that annotates CC$\omega$ terms with sort $U$, lifting them into the non-linear fragment of CLC.

@@

With slight overloading of notation, we define lifting for CC$\omega$ recursively.
\begin{definition}
  \begin{align*}
    \lift{\epsilon}      & = \epsilon                         \\
    \lift{\Gamma, x : A} & = \lift{\Gamma}, x \utype \lift{A}
  \end{align*}
\end{definition}

The following theorems are proved following a similar procedure to proving the soundness of erasure.
\begin{theorem}
  Lift preserves reduction. For any CC$\omega$ term $m$ and $n$, if there is $m \step_{\scriptscriptstyle \text{CC}\omega} n$, then there is $\lift{m} \step_{\scriptscriptstyle \text{CLC}} \lift{n}$.
\end{theorem}

\begin{theorem} \label{lift}
  Lifting. If $\Gamma \vdash m : A$ is a valid judgment in CC$\omega$, then $\lift{\Gamma} \vdash \lift{m} : \lift{A}$ is a valid judgment in CLC.
\end{theorem}

The lifting theorem shows that all valid CC$\omega$ terms are valid CLC terms if the domains of function types are annotated with sort $U$. In practice, these annotations can often be omitted if algorithmic type checking is able to infer them from context.

\section{The Extension to CILC}
Although the core CLC language is extremely expressive, formally shown by Theorem \ref{lift} to be at least as expressive as the predicative CC$\omega$, the complexity of Church-style encodings to do so will quickly become unmanageable. To address this, we extend CLC with rules for defining inductive linear types in the style of CIC.

@@

As shown in Figure \ref{syntax2}, CILC extends CIC with syntax for the introduction and elimination of inductive types. Ind is used during the formation of new inductive types, and Constr is used for introducing their constructors. Case and DCase are non-dependent and dependent case eliminators respectively. CIC-style primitive recursion has been factored out of case eliminators in favor of a Coq-style Fix construct. The flexibility of an independent Fix construct allows for a cleaner formalization of inductive linear type elimination.

\subsection{Arity}
For type $A$ and sort $s$, the judgment $arity(A, s)$ is inductively defined over the structure of $A$ as depicted in Figure \ref{arity}.

@@

\subsection{Strict Positivity}
For type $P$ and variable $X$, the judgment $postive(P, X)$ is inductively defined over the structure of $P$ as depicted in Figure \ref{positive}.

@@

\subsection{Non-Linear Constructor}
For type $C$ and variable $X$, the judgment $\ucons(C, X)$ is inductively defined over the structure of $C$ as shown in Figure \ref{nlconstr}.

@@

\subsection{Linear Constructor}
For type $C$ and variable $X$, the judgment $\lcons(C, X)$ is inductively defined over the structure of $C$ as shown in Figure \ref{lconstr}.

@@

For type $C$ and variable $X$, the judgment $activation(C, X)$ is inductively defined over the structure of $C$ as shown in Figure \ref{activation}. Intuitively, $activation(C,X)$ ensures the linearity of $C$.

@@

\subsection{Introduction Rules}
The formation of new inductive types and their constructors are presented in Figure \ref{ind}.

@@

The rule for non-dependent case elimination is presented in Figure \ref{case}. This elimination rule is non-dependent because the motive $Q$ may not depend on the discriminated term $m$ itself.

@@

Dependent elimination of non-linear inductive objects is presented in Figure \ref{dcase}. This elimination rule is dependent because the motive $Q$ may depend on the discriminated term $m$. For this very reason, $m$ is not allowed to have linear type as doing so may transport restricted variable into types.

@@

\subsection{Fixpoint}
The typing rule of fix-points presented in Figure \ref{fix} is mostly standard. A syntactic guard condition is utilized to conservatively ensure termination. The fix-point term must have non-linear type in a pure context, because recursive calls may cause duplicated usage of restricted variables.

@@

\subsection{Extended Reduction}
As the reduction semantics for the extended language is standard \cite{cic}, we only present the most interesting $\iota$-reduction cases.

@@

\subsection{Linear Connectives}
In Figure \ref{lconnectives}, we demonstrate how the $\otimes$ and $\oplus$ connectives of linear logic could be defined in CILC as inductive linear types. For any term with inductive linear type, linearity will ultimately force the term to be eliminated by a Case expression.

@@

\subsection{Mixed Connectives}
In Figure \ref{mconnectives}, we introduce a $\SigmaL$ connective. $\SigmaL$ is similar to the standard $\Sigma$ type where the type of the second component is dependent on the value of the first component. The key difference between these two type constructors is that $\Sigma$'s second component is non-linear, whereas $\SigmaL$'s second component is linear.

@@

\subsection{Stateful Reasoning}
We first postulate five state axioms. These axioms could be viewed as an interface to a trusted memory allocator. Using these axioms, we will build safe random access arrays from the ground up.
\begin{definition}
  State axioms.
  \begin{itemize}
    \item $\mapsto\ : \mathbb{N} \rightarrow (A : U) \rightarrow L$ \\
          The $\mapsto$ is an infix linear type constructor. It fulfills a role similar to L3's \cite{l3} memory access capability, or Separation Logic's \cite{reynolds02} points-to assertion. Intuitively, a capability $(l \mapsto A)$ is a proof that a term of non-linear type $A$ is currently stored at address $l$.
    \item $\textbf{new} : (A : U) \rightarrow A \rightarrow \SigmaL l:\mathbb{N}.(l \mapsto A)$ \\
          $\new$ is a function that allocates a new memory cell for a term non-linear type $A$ and initializes the cell with its argument. This returns an address $l$ paired with a capability $(l \mapsto A)$.
    \item $\textbf{free} : (A : U) \rightarrow (l:\mathbb{N}) \rightarrow (l \mapsto A) \rightarrow \unit$ \\
          $\free$ is a function that de-allocates the memory cell at location $l$ which is currently in use. The capability $(l \mapsto A)$ is consumed so the memory cell can no longer be accessed.
    \item $\textbf{get} : (A : U) \rightarrow (l : \mathbb{N}) \rightarrow (l \mapsto A) \rightarrow \SigmaL \_: A. (l \mapsto A)$ \\
          $\get$ is a function that retrieves the term stored at address $l$ when given a proof $(l \mapsto A)$ there is indeed a term of type $A$ currently stored there. The type of $\get$'s output is the rather complicated $\SigmaL \_: A. (l \mapsto A)$. This not only returns the term stored at $l$, but also a proof $(l \mapsto A)$ that the memory state has not changed and may be accessed again.
    \item $\textbf{set} : (A\ B : U) \rightarrow (l : \mathbb{N}) \rightarrow B \rightarrow (l \mapsto A) \rightarrow (l \mapsto B)$ \\
          $\set$ has the most complicated type of all the axioms presented here. Intuitively, if address $l$ is allocated and storing some term of type $A$, $\set$ may put a term type $B$ into the the store at address $l$, overwriting the original term. This is the so called "strong update" operation. The consumption of $(l \mapsto A)$ and return of $(l \mapsto B)$ accurately characterize this update process.
  \end{itemize}
\end{definition}
Due to the fact that $(l \mapsto A)$ is an abstract linear type, proofs of this type can not be duplicated and may only be consumed by other state axioms. This is the same non-sharing principle that empowers Separation Logic. Unlike Separation Logic, $(l \mapsto A)$ itself is a first-class object in CILC, meaning that it can be the result of computations (large eliminations), input to computations and storage in data structures, etc.

A random access array is a contiguous region of memory that can be accessed by an address pointing to the start of this region and an offset. In Figure \ref{arrv}, we declare a linear inductive $\arrv$ to track the memory used by the array. $\arrv$ is parameterized by the type $A$ of elements stored in the array and the address $l$ of the initial element. $\arrv$ also takes an additional argument of type $\mathbb{N}$ indicating the length of the array.

>>>>

\def\CComg{{CC}$\omega$}

\begin{document}
\begin{abstract}
  The Calculus of Linear Constructions (CLC) is an extension of the Calculus of Constructions (CC) with linear types. Specifically, CLC extends the predicative \CComg with a hierarchy of linear universes that precisely controls the weakening and contraction of its term level inhabitants. We study the meta-theory of CLC, showing that it is a sound logical framework for reasoning about resources. We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Through examples, we demonstrate that CILC facilitates correct-by-construction imperative programming and lightweight protocol enforcement. We have formalized and proven correct all the reported major results in the Coq Proof Assistant.
\end{abstract}
\maketitle

\section{Introduction}
The Calculus of Constructions (CC) is a dependent type theory introduced by Coquand and Huet in their landmark work~\cite{cc}. In CC, types can depend on terms, allowing one to write precise specifications as types. Today, CC and its extensions CIC~\cite{cic} and ECC~\cite{ecc} lie at the core of popular proof assistants such as Coq~\cite{coq}, Agda~\cite{agda}, Lean~\cite{lean} and others. These theorem provers have found great success in the fields of software verification and constructive mathematics.

However, due to its origins as a logical framework for constructive mathematics, it can be inconvenient for CC to encode and reason about resources. Users of proof assistants based on CC often need to embed external logics such as Separation Logic to provide domain-specific reasoning principles for dealing with resources. We propose an alternative solution: extend CC with linear types.

This paper presents a new linear dependent type system --- the Calculus of Linear Constructions (CLC), which extends the predicative \CComg with linear types. \CComg itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that every resource is used exactly once.

Preexisting approaches to integrating linear types and dependent types often relied on the ! exponential or separate typing judgments for linear and non-linear types. The universe sorts of CLC are enough to completely distinguish between linear and non-linear types within a single typing judgment. As a result, CLC falls closer to its CC roots than prior works in terms of both syntax and semantics.

We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Standard Linear Logic connectives such as $\otimes$ and $\oplus$ can then be defined straightforwardly. And connectives that mix linear and non-linear types can be constructed just as easily in CILC.

Through examples in Section~\ref{examples}, we show some applications of CILC such as construction of safe random access array from the first principles, protocol enforcement through inductively defined session types and in-place update of data structures using continuations.

We have formalized in Coq all the major results reported here and implemented in OCaml a prototype. Additional examples of applications such as provable effect laws and quantitative reasoning are included with our implementation.

\vspace{4pt}
\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item First, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types, enabling direct and precise reasoning about resources.
  \item Second, we study the meta-theory of CLC, showing that it exhibits many desirable properties such as confluence, subject reduction and logical consistency.
  \item Furthermore, we extend CLC with rules to define inductive linear types, forming the Calculus of Inductive Linear Constructions (CILC). This is the first complete presentation of a generalized mechanism for defining inductive linear types.
  \item All major results have been proven correct in Coq. Our development is the first machine checked formalization of a linear dependent type theory. We have also implemented a prototype of CILC in OCaml.
\end{itemize}

@@

The syntax of the core type theory is presented in Figure~\ref{syntax1}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $i$ to quantify over universe levels $0, 1, 2, ...$. $U_i$ and $L_i$ are the universes at level $i$ of non-linear and linear types respectively.
\begin{figure}[h]
  \caption{Correspondence of CLC types and MELL implications}
  \Description{}
  \begin{align}
    (\_ \utype A) \rightarrow B \quad & \Leftrightarrow \quad !(!A \multimap B) \\
    (\_ \ltype A) \rightarrow B \quad & \Leftrightarrow \quad !(A \multimap B)  \\
    (\_ \utype A) \multimap B \quad   & \Leftrightarrow \quad !A \multimap B    \\
    (\_ \ltype A) \multimap B \quad   & \Leftrightarrow \quad A \multimap B
  \end{align}
  \label{correspondence}
\end{figure}

A clear departure of our language from standard presentations of both linear type theory and dependent type theory is the presence of two function types: $(x :_s A) \rightarrow B$ and $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic implicitly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theories. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma~\cite{substitute}. By integrating the exponential directly into universe sorts, we have limited ! to only be canonically usable. This allows us to derive the substitution lemma and construct a direct modeling of CLC in \CComg without requiring any additional machinery for manipulating exponential.

Figure~\ref{correspondence} illustrates the correspondence between CLC functions and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (3) if the co-domain $B$ is dependent on arguments of domain $A$. We will discuss function type formation in Section~\ref{tyformation} in greater detail.

@@

Figure~\ref{universe} illustrates the structure of our universe hierarchy. Each linear universe $L_i$ has $U_{i+1}$ as its type, allowing functions to freely quantify over linear \textit{types}. However, $L_i$ cumulates to $L_{i+1}$. These two parallel threads of cumulativity prevent linear types from being transported to the non-linear universes and subsequently losing track of linearity.

@@

An auxiliary judgment $\pure{\Gamma}$ is defined to assert that a context $\Gamma$ does not contain linear variables. In other words, all variables found in $\pure{\Gamma}$ are annotated of the form $x \utype A$. The full rules for structural judgments are presented in Figure~\ref{structural}.

@@

\subsection{Type Formation} \label{tyformation}
The rules for forming types are presented in Figure~\ref{type}. In CLC, we forbid types from depending on linear terms similarly to~\cite{llf,neel15} for the same reason of avoiding philosophical troubles.

The axiom rules \rname{U-Axiom} and \rname{L-Axiom} are similar to standard \CComg, the main difference being the extra side-condition of judgment $\pure{\Gamma}$. In most presentations of dependent type theories without linear types, the universe axioms are derivable under any well-formed context $\Gamma$. Variables not pertaining to actual proofs could be introduced this way, thus giving rise to the weakening rule. To support linear types, we must restrict weakening to non-linear variables. This justifies the restriction of $\Gamma$ to contain only non-linear variables for \rname{U-Axiom} and \rname{L-Axiom}. From \rname{L-Axiom} we can see that the universe of linear types $L_i$ is an inhabitant of $U_{i+1}$. This is inspired by Krishnaswami et al.'s treatment of linear universes~\cite{neel15}, where linear \textit{types} themselves can be used unrestrictedly.

@@

\subsection{Equality and Reduction} \label{reduction}
The operational semantics and $\beta$-equality of CLC terms are presented in Figure~\ref{red}, all of which are entirely standard.

As we have discussed previously, the elimination of the explicit ! exponential allows CLC to maintain the standard operational semantics of \CComg, whose $\beta$-reductions are very well behaved.

@@

\subsection{Meta Theory} \label{meta}
In this section, we focus our discussion on the properties of CLC. First, we show the type soundness of CLC through the \textit{subject reduction} theorem. Next, through the \textit{linearity} theorem we illustrate how CLC can effectively control usage of resources. Furthermore, we demonstrate that the \textit{promotion} and \textit{dereliction} rules can be encoded as $\eta$-expansions. Finally, we construct a reduction preserving erasure function that maps well-typed CLC terms to well-typed \CComg terms, proving that CLC is strongly normalizing.

All proofs have been formalized in Coq with help from the Autosubst~\cite{autosubst} library. The Coq development is publicly available on the author's Github repository. To the best of our knowledge, this is the first machine checked formalization of a linear dependently type theory. We give a hand written version of the proof in an accompanying technical report.

@@

The proof of confluence is entirely standard using parallel reduction technique~\cite{takahashi}. The presence of linear types does not pose any complications as reductions are untyped.

@@

\subsubsection{Substitution} \label{subst}
Although the substitution lemma is regarded as a boring and bureaucratic result, it is surprisingly hard to design linear typed languages where the substitution lemma is admissible. Much of the difficulty arises during the substitution of arguments containing ! exponential. Perhaps the most famous work detailing the issues of substitution is due to Wadler~\cite{substitute}. He defines intricate syntax and semantics for the unboxing of ! terms, solving the lack of substitute in Abramsky's term calculus.

@@

Notice in Theorem~\ref{nlsubst} of non-linear substitutions, there exists an extra condition $\pure{\Gamma_2}$ which Theorem~\ref{lsubst} of linear substitutions lacks. This condition ensures that linear variables do not occur within the substituted term, thus cannot be duplicated by substitution.

@@

\subsubsection{Strong Normalization}
The strong normalization theorem of CLC is proven by construction of a typing and reduction preserving erasure function from CLC to \CComg. We assume familiarity with \CComg syntax here and define the erasure function as follows.

@@

For the following theorems, we refer to the reductions in CLC as $\step_{\scriptscriptstyle \text{CLC}}$, and the reductions in \CComg as $\step_{\scriptscriptstyle \text{\CComg}}$.

\begin{theorem} \label{preserve}
  Reduction Erasure. For any CLC terms $m$ and $n$, if there is $m \step_{\scriptscriptstyle \text{CLC}} n$, then there is $\erase{m} \step_{\scriptscriptstyle \text{\CComg}} \erase{n}$.
\end{theorem}

\begin{theorem} \label{embedding}
  Embedding. If $\Gamma \vdash m \stype{s} A$ is a valid judgment in CLC, then $\erase{\Gamma} \vdash \erase{m} : \erase{A}$ is a valid judgment in \CComg.
\end{theorem}

If there exists some well-typed CLC term with an infinite sequence of reductions, erasure will embed this term into a well typed \CComg term along with its infinite sequence of reductions by virtue of theorems~\ref{preserve} and~\ref{embedding}. This is contradictory to the strong normalization property of \CComg proven through the Girard-Tait method~\cite{ecc}, so this hypothetical term does not exist in CLC.

@@

\subsubsection{Compatibility}
To show that CLC is compatible with the predicative \CComg, we construct a function that annotates \CComg terms with sort $U$, lifting them into the non-linear fragment of CLC.

@@

With slight overloading of notation, we define lifting for \CComg recursively.
\begin{definition}
  \begin{align*}
    \lift{\epsilon}      & = \epsilon                         \\
    \lift{\Gamma, x : A} & = \lift{\Gamma}, x \utype \lift{A}
  \end{align*}
\end{definition}

The following theorems are proved following a similar procedure to proving the soundness of erasure.
\begin{theorem}
  Lift preserves reduction. For any \CComg term $m$ and $n$, if there is $m \step_{\scriptscriptstyle \text{CC}\omega} n$, then there is $\lift{m} \step_{\scriptscriptstyle \text{CLC}} \lift{n}$.
\end{theorem}

\begin{theorem} \label{lift}
  Lifting. If $\Gamma \vdash m : A$ is a valid judgment in \CComg, then $\lift{\Gamma} \vdash \lift{m} : \lift{A}$ is a valid judgment in CLC.
\end{theorem}

The lifting theorem shows that all valid \CComg terms are valid CLC terms if the domains of function types are annotated with sort $U$. In practice, these annotations can often be omitted if algorithmic type checking is able to infer them from context.

\section{The Extension to CILC}
Although the core CLC language is extremely expressive, formally shown by Theorem~\ref{lift} to be at least as expressive as the predicative \CComg, the complexity of Church-style encodings to do so will quickly become unmanageable. To address this, we extend CLC with rules for defining inductive linear types in the style of CIC.

@@

As shown in Figure~\ref{syntax2}, CILC extends CIC with syntax for the introduction and elimination of inductive types. Ind is used during the formation of new inductive types, and Constr is used for introducing their constructors. Case and DCase are non-dependent and dependent case eliminators respectively. CIC-style primitive recursion has been factored out of case eliminators in favor of a Coq-style Fix construct. The flexibility of an independent Fix construct allows for a cleaner formalization of inductive linear type elimination.

\subsection{Arity}
For type $A$ and sort $s$, the judgment $arity(A, s)$ is inductively defined over the structure of $A$ as depicted in Figure~\ref{arity}.

@@

\subsection{Strict Positivity}
For type $P$ and variable $X$, the judgment $postive(P, X)$ is inductively defined over the structure of $P$ as depicted in Figure~\ref{positive}.

@@

\subsection{Non-Linear Constructor}
For type $C$ and variable $X$, the judgment $\ucons(C, X)$ is inductively defined over the structure of $C$ as shown in Figure~\ref{nlconstr}.

@@

\subsection{Linear Constructor}
For type $C$ and variable $X$, the judgment $\lcons(C, X)$ is inductively defined over the structure of $C$ as shown in Figure~\ref{lconstr}.

@@

For type $C$ and variable $X$, the judgment $activation(C, X)$ is inductively defined over the structure of $C$ as shown in Figure~\ref{activation}. Intuitively, $activation(C,X)$ ensures the linearity of $C$.

@@

\subsection{Introduction Rules}
The formation of new inductive types and their constructors are presented in Figure~\ref{ind}.

@@

The rule for non-dependent case elimination is presented in Figure~\ref{case}. This elimination rule is non-dependent because the motive $Q$ may not depend on the discriminated term $m$ itself.

@@

Dependent elimination of non-linear inductive objects is presented in Figure~\ref{dcase}. This elimination rule is dependent because the motive $Q$ may depend on the discriminated term $m$. For this very reason, $m$ is not allowed to have linear type as doing so may transport restricted variable into types.

@@

\subsection{Fixpoint}
The typing rule of fix-points presented in Figure~\ref{fix} is mostly standard. A syntactic guard condition is utilized to conservatively ensure termination. The fix-point term must have non-linear type in a pure context, because recursive calls may cause duplicated usage of restricted variables.

@@

\subsection{Extended Reduction}
As the reduction semantics for the extended language is standard~\cite{cic}, we only present the most interesting $\iota$-reduction cases.

@@

\subsection{Linear Connectives}
In Figure~\ref{lconnectives}, we demonstrate how the $\otimes$ and $\oplus$ connectives of linear logic could be defined in CILC as inductive linear types. For any term with inductive linear type, linearity will ultimately force the term to be eliminated by a Case expression.

@@

\subsection{Mixed Connectives}
In Figure~\ref{mconnectives}, we introduce a $\SigmaL$ connective. $\SigmaL$ is similar to the standard $\Sigma$ type where the type of the second component is dependent on the value of the first component. The key difference between these two type constructors is that $\Sigma$'s second component is non-linear, whereas $\SigmaL$'s second component is linear.

@@

\subsection{Stateful Reasoning}
We first postulate five state axioms. These axioms could be viewed as an interface to a trusted memory allocator. Using these axioms, we will build safe random access arrays from the ground up.
\begin{definition}
  State axioms.
  \begin{itemize}
    \item $\mapsto\ : \mathbb{N} \rightarrow (A : U) \rightarrow L$ \\
          The $\mapsto$ is an infix linear type constructor. It fulfills a role similar to L3's~\cite{l3} memory access capability, or Separation Logic's~\cite{reynolds02} points-to assertion. Intuitively, a capability $(l \mapsto A)$ is a proof that a term of non-linear type $A$ is currently stored at address $l$.
    \item $\textbf{new} : (A : U) \rightarrow A \rightarrow \SigmaL l:\mathbb{N}.(l \mapsto A)$ \\
          $\new$ is a function that allocates a new memory cell for a term non-linear type $A$ and initializes the cell with its argument. This returns an address $l$ paired with a capability $(l \mapsto A)$.
    \item $\textbf{free} : (A : U) \rightarrow (l:\mathbb{N}) \rightarrow (l \mapsto A) \rightarrow \unit$ \\
          $\free$ is a function that de-allocates the memory cell at location $l$ which is currently in use. The capability $(l \mapsto A)$ is consumed so the memory cell can no longer be accessed.
    \item $\textbf{get} : (A : U) \rightarrow (l : \mathbb{N}) \rightarrow (l \mapsto A) \rightarrow \SigmaL \_: A. (l \mapsto A)$ \\
          $\get$ is a function that retrieves the term stored at address $l$ when given a proof $(l \mapsto A)$ there is indeed a term of type $A$ currently stored there. The type of $\get$'s output is the rather complicated $\SigmaL \_: A. (l \mapsto A)$. This not only returns the term stored at $l$, but also a proof $(l \mapsto A)$ that the memory state has not changed and may be accessed again.
    \item $\textbf{set} : (A\ B : U) \rightarrow (l : \mathbb{N}) \rightarrow B \rightarrow (l \mapsto A) \rightarrow (l \mapsto B)$ \\
          $\set$ has the most complicated type of all the axioms presented here. Intuitively, if address $l$ is allocated and storing some term of type $A$, $\set$ may put a term type $B$ into the the store at address $l$, overwriting the original term. This is the so called "strong update" operation. The consumption of $(l \mapsto A)$ and return of $(l \mapsto B)$ accurately characterize this update process.
  \end{itemize}
\end{definition}
Due to the fact that $(l \mapsto A)$ is an abstract linear type, proofs of this type can not be duplicated and may only be consumed by other state axioms. This is the same non-sharing principle that empowers Separation Logic. Unlike Separation Logic, $(l \mapsto A)$ itself is a first-class object in CILC, meaning that it can be the result of computations (large eliminations), input to computations and storage in data structures, etc.

A random access array is a contiguous region of memory that can be accessed by an address pointing to the start of this region and an offset. In Figure~\ref{arrv}, we declare a linear inductive $\arrv$ to track the memory used by the array. $\arrv$ is parameterized by the type $A$ of elements stored in the array and the address $l$ of the initial element. $\arrv$ also takes an additional argument of type $\mathbb{N}$ indicating the length of the array.
